Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-18 cordillera-cycle.tex
Polish conclusions.

The possibly of a widespread late-glacial in
the North American Cordilleran has been debated for long. On the southern tip
of the Coast Mountains, radiocarbon-dated frontal moraines indicate multiple
stages of glacial re-advance in the Fraser and Squamish valleys, of which
one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. In the Finlay River area, northern Rocky Mountains,
sharp-crested moraines show an abundant remobilization of late-glacial
sediments by alpine glaciers in interaction with the main body of the decaying
ice sheet \citep{Lakeman.etal.2008}. Additional clues for a late-glacial
re-advance include alpine glacier moraines in the eastern Coast Mountains,
Rocky Mountains and Columbia Mountains \citep{Menounos.etal.2008}. Although
further work is needed to constrain the timing of the late-glacial re-advance,
to seize its extent and to identify potential climatic triggers
\citep{Menounos.etal.2008}, it is interesting to note that the simulation
driven by the GRIP record produces a late-glacial re-advance in the Coast
Mountains, Rocky Mountains and Finlay River area, corresponding to where it
has been identified in the geologic record (Fig.~\ref{fig:deglac}, left panel),
whereas the EPICA-driven simulation produces a nearly-continuous deglaciation
with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel).

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glaciation maxima during
MIS~4 (61.9--55.4\,kyr) and MIS~2 (29.5--16.9\,kyr), two periods
corresponding to documented extensive palaeo-glaciations. This result is
independent of the choice of palaeo-temperature record used to approximate the
past evolution of climate forcing, and
can be seen as a first-order agreement between the model and geomorphological
evidence. However, the timing of the two glaciation peaks depends sensitively
on which record
is used to drive the model. The timing of the last glacial maximum is best
reproduced by the Antarctic record, and occurs too early in all simulations
that are driven by other records. The mismatch is greatest when using
oceanic records from the Pacific Northwest, which are affected by the
weakening of the California current during the last glacial maximum.

In all simulations presented here, ice cover is limited to disjoint mountain ice
caps during most of the glacial cycle. This agrees with earlier interpretations
of the geological evidence preceding the last glacial maximum. However, our
simulations produce persistent ice cover over the Skeena mountains during
the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
absent, the Skeena ice cap appears to be fed by the eastwards precipitation
intrusion through a topographic window in the Coast Mountains. The ice cap acts
as a nucleation centre at the onset of the last glacial maximum re-advance, and
may have contributed to the spectacular glacial imprint of the Skeena Mountains
erosional landscape.

Concerning the deglaciation phase, when most geological evidence is available,
none of the palaeo-temperature records used produces an ideal agreement between
model results and geological evidence. The timing of the last glacial maximum
and early deglaciation is generally best reproduced by forcing the model with
Antarctic records. However, the Younger Dryas re-advance is only reproduced
using temperature forcing from the GRIP record at places where it has been
documented. Nonetheless, the patterns of deglaciation are consistent between
both simulations driven by the GRIP and EPICA record. They show a rapid
deglaciation of the southern half of the ice
sheet, followed by unzipping of the Liard Lowland, and concentric retreat of
the ice margin towards the last palaeo-ice cover in the Skeena Mountains
during the early Holocene (10.9--9.5\,kyr).

One must keep in mind, however, that these results are only accurate for our
choices of ice-sheet model (PISM), surface mass balance model (PDD) and climate
forcing. Most importantly, our simplistic
palaeo-climate forcing does not include precipitation corrections in response
to the presence of an ice cover, likely leading to overestimated glacial
extent and volume. Nevertheless, our results identify that the largely
understudied Skeena Mountains are a key area to understanding glacial dynamics
of the Cordilleran ice sheet, highlighting the need for further geological
investigation of this region.

>>>>

The possibly of a widespread late-glacial in
the North American Cordilleran has been debated for long. On the southern tip
of the Coast Mountains, radiocarbon-dated frontal moraines indicate multiple
stages of glacial re-advance in the Fraser and Squamish valleys, of which
one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. In the Finlay River area, northern Rocky Mountains,
sharp-crested moraines show an abundant remobilization of late-glacial
sediments by alpine glaciers in interaction with the main body of the decaying
ice sheet \citep{Lakeman.etal.2008}. Additional clues for a late-glacial
re-advance include alpine glacier moraines in the eastern Coast Mountains,
Rocky Mountains and Columbia Mountains \citep{Menounos.etal.2008}. Although
further work is needed to constrain the timing of the late-glacial re-advance,
to seize its extent and to identify potential climatic triggers
\citep{Menounos.etal.2008}, it is interesting to note that the simulation
driven by the GRIP record produces a late-glacial re-advance in the Coast
Mountains, Rocky Mountains and Finlay River area, corresponding to where it
has been identified in the geologic record (Fig.~\ref{fig:deglac}, left panel),
whereas the EPICA-driven simulation produces a nearly-continuous deglaciation
with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel)%
\footnote{Should we mention our work in progress with Brian and/or the Tuya
    Lake moraine?}.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glaciation maxima during
MIS~4 (61.9--55.4\,kyr) and MIS~2 (29.5--16.9\,kyr), two periods
corresponding to documented extensive glaciations. This result is
independent of the choice of palaeo-temperature record used to approximate past
climate evolution, and thus
can be seen as a first-order agreement between the model and geological
evidence. However, the timing of the two glaciation peaks depends sensitively
on which record
is used to drive the model. The timing of the last glacial maximum is best
reproduced by the Antarctic record, and occurs too early in all simulations
that are driven by other records. The mismatch is greatest when using
oceanic records from the Pacific Northwest, which are affected by the
weakening of the California current during the last glacial maximum.

In all simulations presented here, ice cover is limited to disjoint mountain ice
caps during most of the glacial cycle, confirming previous inferences from
geological evidence preceding the last glacial maximum. However, our
simulations produce persistent ice cover on the Skeena mountains during
the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
absent, the Skeena ice cap appears to be fed by the eastwards precipitation
intrusion through a topographic window in the Coast Mountains. The ice cap acts
as a nucleation centre at the onset of the last glacial maximum re-advance, and
may have contributed to the spectacular glacial imprint of the Skeena Mountains
erosional landscape.

During deglaciation, none of the palaeo-temperature records used produces an
ideal agreement between model results and geological evidence. Although
the EPICA record yields to more realistic timing of the last glacial maximum
and early deglaciation, only the GRIP record produces a late-glacial re-advance
at places where it has been documented in the literature. Nonetheless, the
general patterns of deglaciation are consistent between both simulations driven
by the GRIP and EPICA record, and show a rapid deglaciation of the southern
half of the ice including rapid northwards retreat through the Interior
Plateau of central British Comumbia, followed by an opening of the Liard
Lowland, and final ice cap retreat of the ice margin towards the Selwyn and,
then, the Skeena Mountains, which host the last remnant of the ice sheet
during the early Holocene (10.9--9.5\,kyr).

These results strongly depend on our choices of ice-sheet model (PISM), surface
mass balance model (PDD) and climate forcing. Most importantly, our simplistic
palaeo-climate forcing does not include precipitation corrections in response
to the presence of an ice cover, potentially leading to overestimated glacial
extent and volume in continental regions. Nevertheless, our results identify
the glacially-scrubbed Skeena Mountains as a central, key area to understanding
glacial dynamics of the Cordilleran ice sheet, highlighting the need for
further geological investigation of this region.
